Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-06-18 doc/manuscript_hkk_drought_sensitivity.Rmd
text snipping
current count - 5106

**Number of Words in Abstract:** 150  
**Number of Words in Main Text:** 5200  
**Number of References:** 46  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

@@

Intense droughts are becoming more frequent, affecting the demography, diversity, and  carbon (C) cycling of forests globally [@ipcc_climate_2023]. 
However, our ability to predict drought impacts on tree woody growth and ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_cryptic_2019] [@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

@@

Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
This is a 50-hectare plot in dry seasonal evergreen forest, with a mean annual precipitation of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_. 
For analysis, we focused on 30 of the most abundant species in the plot.   
<!-- Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].   -->

@@

Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year for each tree. 
We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
We used the raw window size measurements combined with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
We used a direct measurement of $d_{1}$ at the time of installation if it was available.
If the tape measurement at the first window size measurement was not available for any band, we excluded window size measurements until the first available diameter measurement. 
We then calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).

To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$.
We also excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations). 
Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for each tape measure is 1 mm, error for growth is at least 2 mm. 
In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
However, tape measurements to flag likely errors in the dendrometer dataset. 
We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these annualised increments calculated from tape measurements. 
Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.

@@

<!-- We identified 2010 and 2015 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.  -->
We identified 2010, 2015 and 2020 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses. 
We calculated Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
Defining drought as periods with long-term SPEI < -1 resulted in identification of 2010, 2015 and 2020 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig. `r fig_droughts`) and plot-level growth declines in the timeseries (Fig. `r fig_drought_responses`). 
Among these, 2010 and 2015 were El Niño years; 2010 was a moderate El Niño event while 2015 was a very strong El Niño event (Oceanic Niño Index @bamston_documentation_1997). 
The drought in 2020 was not associated with an El Niño event.
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
<!-- We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.     -->

<!-- To characterise the two focal droughts, we used data from a weather station near the plot and publicly available climate data for the region.  -->
<!-- We used daily weather data on maximum temperate, relative humidity and total precipitation collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from HKK, and calculated an additional variables vapour pressure deficit (VPD, @raesch_pvldcurve_2020).  -->
To characterise the three droughts, we used daily weather data obtained from publicly available data sources. 
We used daily weather data obtained from remotely sensed data source CHIRPS [@funk_climate_2015] for daily total precipitation and ERA5Land [@muñoz-sabater_era5land_2021] to calculate daily maximum temperature and vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30 day periods across the whole timeseries.
Finally, we calculated daily rolling anomalies for each variable for the drought years from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$

@@

We first  calculated the mean annual increment $\bar{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$<!-- as $\frac{\[ 
  \sum_{t}^{}{ainc_{i,t}} 
  \]}{\sum{t}}$-->.
We then calculated sensitivity for each tree $i$ for each drought year $y$ as 
$$
sensitivity_{i, t=y} = \frac{ainc_{i, t=y} - \bar{a.inc_{i}}}{\bar{a.inc_{i}}}
$$
By scaling the difference in increment to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species.

@@

Across all individuals in the plot, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location. 
We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment.
We used the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to fill and breach this layer, calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location. 
We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
We estimated crown exposure for each tree at each census using crown illumination index (CII), an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census.

@@

We analysed data for each drought year by fitting separate models. 
We fit alternative models for sensitivity including a null model with just species random effects, a model including only fixed effects of species deciduousness and habitat wetness and a model including size (DBH), light environment (CII) and habitat wetness as well as species random effects on all these.   
We first modelled sensitivity as a simple random intercept model

@@

The three droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
The 2010 drought peaked between February and May with meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
<!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period. 
<!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought. 
In 2020, meteorological anomalied were pronounced both in the dry and wet seasons, including maximum temperature (max anomaly **> 2** SD) and VPD (max anomaly **> 2** SD). 
1- to 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
12-month SPEI showed severe to moderate drought conditions from January to August, spanning both dry and wet seasons.

@@

Species median sensitivities to the three droughts were not correlated with each other although on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020 was r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for  2010 and 2020 r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).

@@

Crown exposure was associated with more negative sensitivities in all three droughts.
In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`). 
Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig `r fig_dag`b).

>>>>

**Number of Words in Abstract:** 150  
**Number of Words in Main Text:** 5106  
**Number of References:** 54  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

@@

Intense droughts are becoming more frequent, affecting the demography, diversity, and  carbon (C) cycling of forests globally [@ipcc_climate_2023]. 
However, our ability to predict drought impacts on tree woody growth and ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove more vulnerable in a different drought. 
The complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of the climatic anomalies [@slette_how_2019].
<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_cryptic_2019] [@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

@@

Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_. 
<!-- For analysis, we focused on 30 of the most abundant species in the plot.    -->
<!-- Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].   -->

@@

Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year and tree. 
We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
We used window size measurements combined with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
We used a direct measurement of $d_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).

<!-- To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$. -->
We excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations). 
Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for each measurement is 1 mm, error for growth is 2 mm. 
In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
However, we used tape measurements to flag likely errors in the dendrometer dataset. 
We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement. 
Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.

@@

<!-- We identified 2010 and 2015 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.  -->
We identified 2010, 2015 and 2020 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses. 
We calculated Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
Defining drought as periods with long-term SPEI < -1 resulted in identification of 2010, 2015 and 2020 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig. `r fig_droughts`) and plot-level growth declines in the timeseries (Fig. `r fig_drought_responses`). 
Among these, 2010 and 2015 were El Niño years; 2010 was a moderate El Niño event while 2015 was a very strong El Niño event (Oceanic Niño Index @bamston_documentation_1997). 
<!-- The drought in 2020 was not associated with an El Niño event. -->
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
<!-- We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.     -->

<!-- To characterise the two focal droughts, we used data from a weather station near the plot and publicly available climate data for the region.  -->
<!-- We used daily weather data on maximum temperate, relative humidity and total precipitation collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from HKK, and calculated an additional variables vapour pressure deficit (VPD, @raesch_pvldcurve_2020).  -->
To characterise the three droughts, we used daily weather data obtained from publicly available data sources. 
We used data from CHIRPS [@funk_climate_2015] for daily total precipitation and ERA5Land [@muñoz-sabater_era5land_2021] to calculate daily maximum temperature and vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30-day periods across the whole timeseries.
Finally, we calculated daily rolling anomalies for each variable for the drought years from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$

@@

We first  calculated the mean annual increment $\bar{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$.
We then calculated sensitivity for each tree $i$ for each drought year $y$ as 
$$
sensitivity_{i, t=y} = \frac{ainc_{i, t=y} - \bar{a.inc_{i}}}{\bar{a.inc_{i}}}
$$
By scaling the increment deviation to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species.

@@

Across all individuals, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location. 
We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment and the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
<!-- As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location.  -->
For an alternate estimate of wetness (Topographic Position Index) and associated results, see supplementary material. 
We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
We estimated crown exposure for each tree at each census using crown illumination index (CII), an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census. 
We centred and scaled TWI and DBH across the range of observed values.

@@

We analysed tree responses to individual droughts by fitting separate models. 
We fit alternative models for sensitivity including a null model with just species random effects, a model including only fixed effects of species deciduousness and habitat wetness and a model including size (DBH), light environment (CII) and habitat wetness as well as species random effects on all these. 
We first modelled sensitivity as a simple random intercept model

@@

The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
The 2010 drought showed 30-day-mean meteorological anomalies between February and May including lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD), including in the wet season.
<!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
<!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
<!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
1- to 12-month SPEI values crossed minima -2, lowest in the dendroband census period and indicative of severe drought. 
In 2020, meteorological anomalies were pronounced both in the dry and wet seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
3- and 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry and wet seasons.

@@

Species median sensitivities to the three droughts were not correlated with each other, although on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020 was r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for  2010 and 2020 r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).

@@

Crown exposure was associated with greater growth declines in all three droughts.
In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`). 
Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig `r fig_dag`b).
